---
title: Analysis of financial real-world Data sets
output: html_document

---

# Introduction
This report provides a comprehensive analysis of real-world financial data sets, focusing on the S&P 500 and Bitcoin. The purpose is to uncover underlying patterns, assess market volatility, and evaluate the correlation between these two prominent financial indicators.

# Task 1

<p style="color: blue;">Descriptive Statistics: Calculate, for example, the mean, median, mode, range, standard deviation, etc., of both the S&P data and Bitcoin data. Compare the measures of central tendency and variability between the two datasets.</p>


```{r}
# Install packages
install.packages("zoo")
library(zoo)
```

```{r}
# Install ggpubr
install.packages("ggpubr")

# Load the ggpubr package
library(ggpubr)
```

```{r}
install.packages("modeest")
```

```{r}
install.packages("knitr")
```

```{r}
# Load the necessary packages
library(readr)
library(dplyr)
library(ggplot2)
library(zoo)
library(modeest)
library(knitr)

# Load the data
sp_data <- read_csv("S&P 500.csv")
bitcoin_data <- read_csv("BTC-USD.csv")

# Check date formats and data types again
print(str(sp_data))
print(str(bitcoin_data))


# Inspect the first few rows of the datasets
head(sp_data)
head(bitcoin_data)
```

```{r}
# Define a function to calculate all relevant statistics
calculate_stats <- function(data, price_col) {
  stats <- data %>%
    summarise(
      Mean = mean({{price_col}}, na.rm = TRUE),
      Median = median({{price_col}}, na.rm = TRUE),
      Mode = as.numeric(mfv({{price_col}})[1]),  # Take the first mode value if multiple
      Range = max({{price_col}}, na.rm = TRUE) - min({{price_col}}, na.rm = TRUE),
      Standard_Deviation = sd({{price_col}}, na.rm = TRUE),
      Variance = var({{price_col}}, na.rm = TRUE)
    )
  return(stats)
}

# Calculate statistics for S&P 500
sp_stats <- calculate_stats(sp_data, Price)

# Calculate statistics for Bitcoin
bitcoin_stats <- calculate_stats(bitcoin_data, price)

# Print the results in a more visual way using knitr::kable()
cat("S&P 500 Descriptive Statistics\n")
print(knitr::kable(sp_stats))

cat("\nBitcoin Descriptive Statistics\n")
print(knitr::kable(bitcoin_stats))
```

## Central Tendency (Mean, Median, Mode)
### Mean:
S&P 500: 3867.881
Bitcoin: 27097.76
Interpretation: Bitcoin's average price is significantly higher than that of the S&P 500, suggesting that on average, Bitcoin's trading price is higher over the observed period.
### Median:
S&P 500: 3971.18
Bitcoin: 25576.39
Interpretation: Similar to the mean, the median price of Bitcoin is much higher than the S&P 500. The median being lower than the mean in Bitcoin’s case suggests a right-skewed distribution, where higher prices pull the mean upwards more than the median.
### Mode:
S&P 500: 2926.46
Bitcoin: 4879.878
Interpretation: The mode, or most frequently occurring price, is also much lower for the S&P 500 compared to Bitcoin. The mode's position much below the mean and median for Bitcoin again suggests a skewed distribution, possibly with significant price spikes.

## Variability (Range, Standard Deviation, Variance)
### Range:
S&P 500: 3016.95
Bitcoin: 68203.62
Interpretation: Bitcoin shows a much wider range than the S&P 500, indicating that Bitcoin’s prices fluctuated over a much larger scale. This points to higher volatility in the Bitcoin market.
### Standard Deviation:
S&P 500: 643.1942
Bitcoin: 16711.71
Interpretation: Bitcoin's standard deviation is substantially higher than that of the S&P 500, confirming that Bitcoin prices are more spread out from the mean, which is indicative of higher volatility.
### Variance:
S&P 500: 413698.8
Bitcoin: 279281159
Interpretation: The variance, like standard deviation, is much higher for Bitcoin. A higher variance indicates a wider dispersion of price values around the mean, reinforcing the notion of Bitcoin as a more volatile asset.

# Task 2
<p style="color: blue;">By drawing a plot, showing the trend or pattern of S&P data and Bitcoin data over five year</p>

```{r}
# Ensure the Date columns are in Date format
sp_data$Date <- as.Date(sp_data$Date)
bitcoin_data$Date <- as.Date(bitcoin_data$Date)
```

```{r}
# Plotting S&P 500 trend
ggplot(sp_data, aes(x = Date, y = Price)) +
  geom_line(color = 'blue') +
  labs(title = "Trend of S&P 500 Over Five Years", x = "Date", y = "Price") +
  theme_minimal()


```

```{r}
# Plotting Bitcoin trend
ggplot(bitcoin_data, aes(x = Date, y = price)) +
  geom_line(color = 'red') +
  labs(title = "Trend of Bitcoin Over Five Years", x = "Date", y = "price") +
  theme_minimal()
```

The S&P 500 plot shows whether there has been a steady increase, decrease, or fluctuation over the period. Generally, stock indexes like the S&P 500  show upward trends with occasional dips.
The Bitcoin plot shows more volatility with potentially higher peaks and deeper troughs, reflecting the high-risk high-return nature of cryptocurrencies.

```{r}
# Merge data based on Date
combined_data <- merge(sp_data, bitcoin_data, by = "Date")

# Ensure that data frame is sorted by date to correctly calculate rolling windows
combined_data <- combined_data %>% arrange(Date)

# Calculate the rolling correlation correctly
combined_data$RollingCorrelation <- rollapply(width = 182,
                                              data = combined_data[, c("Price", "price")],
                                              FUN = function(x) {
                                                # Ensure the input to cor() is correctly formatted
                                                if (ncol(x) == 2) {
                                                  cor(x[, 1], x[, 2], use = "complete.obs")
                                                } else {
                                                  NA  # Return NA if there's any issue with the data input
                                                }
                                              },
                                              by.column = FALSE,
                                              align = "right",
                                              fill = NA)

# Plotting the rolling correlation
ggplot(combined_data, aes(x = Date, y = RollingCorrelation)) +
  geom_line(color = 'purple') +
  labs(title = "Rolling Correlation Between S&P 500 and Bitcoin Over Five Years",
       x = "Date", y = "Correlation") +
  theme_minimal()
```

# Task 3
<p style="color: blue;">Compute the correlation coefficient between S&P data and Bitcoin data to explore if there's any relationship between the two</p>

```{r}
# Check for any NA values and remove rows with NA values if any
combined_data <- na.omit(combined_data)

# Ensure that both columns are numeric
#combined_data$Price_sp <- as.numeric(combined_data$Price_sp)
#combined_data$Price_btc <- as.numeric(combined_data$Price_btc)

# Check the head of the merged dataset to confirm successful merge
print(head(combined_data))
print(nrow(combined_data))
```

```{r}
print(paste("Length of S&P 500 Prices:", length(combined_data$Price)))
print(paste("Length of Bitcoin Prices:", length(combined_data$price)))
print(paste("Type of S&P 500 Prices:", class(combined_data$Price)))
print(paste("Type of Bitcoin Prices:", class(combined_data$price)))
```

```{r}
# Compute the correlation coefficient
correlation_coefficient <- cor(combined_data$Price, combined_data$price, use = "complete.obs")
print(paste("Correlation Coefficient between S&P 500 and Bitcoin:", correlation_coefficient))

```

```{r}
# Scatter plot of S&P 500 prices vs. Bitcoin prices
ggplot(combined_data, aes(x = Price, y = price)) +
  geom_point(alpha = 0.6, color = 'blue') +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Scatter Plot of S&P 500 vs. Bitcoin Prices",
       x = "S&P 500 Price",
       y = "Bitcoin Price") +
  theme_minimal()

```

## Correlation Coefficient:
A value close to 1 indicates a strong positive correlation, meaning as S&P 500 prices go up, Bitcoin prices also tend to go up.
A value close to -1 indicates a strong negative correlation, meaning as S&P 500 prices go up, Bitcoin prices tend to go down.
A value close to 0 suggests a weak or no linear relationship.

Hence it is weak or no linear relationship.

## Scatter Plot Analysis:
The geom_smooth(method = "lm") line (red line) shows the best fit linear trend line through the data, providing a visual aid to understand the correlation.

# Task 4
<p style="color: blue;">Assess whether the S&P data and Bitcoin data follow a normal distribution.</p>

Assess whether the S&P data and Bitcoin data follow a normal distribution. Analyze each dataset separately and interpret the findings

```{r}
# Q-Q plot for S&P 500
ggqqplot(sp_data$Price, title = "Q-Q Plot for S&P 500")

# Q-Q plot for Bitcoin
ggqqplot(bitcoin_data$price, title = "Q-Q Plot for Bitcoin")
```

## Bitcoin Q-Q Plot Interpretation:
*   The plot shows that the data points deviate significantly from the line, especially at the ends, which indicates that the distribution has heavier tails than a normal distribution.
*   There is a pronounced curve away from the normal line, suggesting that the Bitcoin prices have a skewed distribution with a lot of data concentrated on one side of the mean and a long tail on the other side.
* The S-shaped curve indicates that the distribution has more extreme values (both high and low) than what would be expected in a normal distribution, which is characteristic of "fat tails".


## S&P 500 Q-Q Plot Interpretation:
* Similar to the Bitcoin plot, there is a noticeable deviation from the line, particularly at the tails, suggesting that the S&P 500 data also have heavier tails than a normal distribution.

* The curvature, especially on the upper tail, indicates that the S&P 500 data is right-skewed, with a long tail of higher values.

* The middle portion of the plot, which is relatively closer to the line, indicates that a large portion of the data does follow a pattern similar to a normal distribution, but the distribution overall is not normal due to the deviations at the tails.

# Conclusion
This report detailed the statistical analysis and visual trends of the S&P 500 and Bitcoin, revealing key insights about their market behaviors, volatility, and interconnectivity. The findings indicate significant differences in market dynamics, which are crucial for investors and policymakers.
